
\section*{Introduction}

\epigraph{\singlespacing In short, \emph{both} state capacity and politics must be studied if we are to explain state performance -- especially in the developing world.}{\singlespacing Cente\~{n}o, Kohli and Yashar, \emph{States in the Developing World}}

\doublespacing

Citizens across the world rely on governments to provide key public services. Whether it be to administer a vaccine or teach the alphabet, bureaucrats are responsible for fulfilling this government mandate. These public officials span all levels of government: from the frontline, to management and the highest levels of public office, bureaucrats play a crucial role in shaping who receives public services and how these are administered. Under the broad umbrella of state capacity, I sought to understand how to make bureaucracies work better. I focused on why these actors behave the way they do, all the while raising questions regarding the political dynamics in which they are embedded.

In this dissertation, I explore the political administration of bureaucracies. Using Brazilian local governments as a case study, I analyze how political institutions and actors shape well-known empirical phenomena such as patronage and corruption. The goal was to understand how and why bureaucrats and politicians alike choose to behave the way they do. In interviews, formal modeling and exploratory data analyses, I have strived to illustrate their choices and incentive structures, both individual and institutional, that lead them to act ways that are often counterproductive to the quality of management of public services. It is my hope that shedding light on how these decisions take place can inform the design of better policies to reshape them.

In the first chapter, I explore how local governance institutions shape patronage dynamics in the educational sector. Combining qualitative fieldwork evidence with large scale administrative data, I argue that patronage is driven by local politician's need to coopt the local legislature. In particular, I find that, consistent across different model specifications, the stronger support a mayor has in the legislature, the less patronage occurs. A formal model on buying supermajorities provides a theoretical framework to understand this finding. Additionally, I show that patronage appointments have knock-on effects on educational outcomes, with lower test scores in schools exposed to staff turnover.

In the second chapter, coauthored with Romain Ferrali, we focus squarely on bureaucrat's decisions to engage in corruption. We combine randomized anti-corruption audits with structural estimation to identify the conditions under which auditing technologies can reduce corruption in Brazilian local governments. We find that anti-corruption audits have only limited effects on career choices for high-level bureaucrats. In particular, in municipalities found to be corrupt and exposed to the audits there is no evidence of changes in dismissals or departures of bureaucrats. To understand this finding, we model bureaucrats' choice to engage in corruption and apply our model to the data through structural estimation. Counterfactual analysis identifies strong complementarities between the components of the audit, suggesting that pulling all levers may be more effective at reducing bureaucratic corruption.

The last chapter focuses on understanding the breadth of partisan affiliation in local bureaucrats. Joining micro-level data on party membership with employment histories of all formal sector workers, I shed light on the career paths and characteristics of party members across private and public sectors. Career data suggests that party members who enter the bureaucracy are primarily selected from the high-income tail of the distribution in the private sector. Once hired, party members concentrate in the upper levels of management in the public sector, receiving contract tenure, higher wages and staying for longer than non-partisan counterparts. Finally, I provide evidence that patronage can be used to incentivize wealthy donors to contribute to local parties, raising questions regarding party building and finance, as well as representation in developing contexts.

Understanding how to make bureaucracies work better is necessarily an unfinished agenda. The multiplicity of bureaucratic institutions, actors, and government structures simply defy a unique answer. That is our asymptotic frontier. However, I remain convinced that a combination of methods can shed light on this complex, but structured, institutions. In fact, it may be the only principled way to move forward and approximate us to truths. I hope that the structure of this dissertation reflects this intellectual humility: states need to be unpacked, and meso-level theories that bring together substantive concerns, theoretical modeling and rich data -- both qualitative and quantitative -- can provide us with tentative but grounded answers to pressing questions regarding state capacity. 

Much has been said, and much remains to be done. It is my hope that the projects I have embarked on in my research career at Princeton will leave a trail behind that other scholars can explore. All the data and replication files used for this dissertation are open source and can be found at my GitHub page \href{https://github.com/galileukim}{here}. As I have learned from others, I hope that other scholars can learn from my advances, avoid my shortcomings, and push even further ahead. My journey does not end here though! I recently joined the Bureaucracy Lab at the World Bank, as well as the Knowledge, Innovations and Communications Team at the IDB. I am partnering with the Brazilian National Public Administration Department for some research projects on bureaucracies. There is much to be done.

\newpage

\section*{Acknowledgments}

Science, especially a social one, is a collective endeavor. Fortunately, at Princeton I have not been only able to stand on the shoulder of giants, but learn from them. 

I am grateful to my dissertation commitee for their mentorship and guidance in this journey. Deborah Yashar, my chair, has taught me how to become a scholar, a lifelong lesson I will cherish dearly. Deborah's kind support and constructive insights have led me to intellectual heights I could never reach alone. Matias Iaryczower has taught me how to take ownership and responsibility for my work, and I am immensely grateful for that. From the classroom to coauthorship, I have learned so much from you. Gracias Mati. Guadalupe Tu\~{n}on has generously advised me, in the final stretch, to take a step back, think clearly and walk forward. Working together on these research projects has been a delight. Thank you all.

Friends and colleagues have reminded me constantly that it does not matter how far we go, we can always push forward together. I am grateful to Jos\'{e} Mar\'{i}a Rodriguez Valadez, Will Horne, Dan Gibbs, Beatriz Barros, James Mao, Guilherme Duarte, Elsa Voytas, Michael Kistner, Naoki Egami, and Carolyn Barnett for their friendship and solidarity in our collective journey. Romain Ferrali, one of the coauthors for a chapter in my dissertation, has encouraged me to become a better researcher, developer and professional, all to a jazzy tune. Matheus Hardt has joined forces with me tackling the Brazilian educational data and later, in our data science forays for the Brazilian federal government. Members of the Latin American Politics Workshop, Brazil Lab, Program for Latin American Studies and Political Economy Colloquium have pushed my research agenda with their constructive feedback and curiosity. Thank you.

Fieldwork in Brazil would not have been possible without the generous help of Norman Gall and the team at the Instituto Fernando Braudell. They have received me with open arms, and I have learned tremendously from their generosity. Sandra Leite and the team at SEDUC, UDIME and COPEM have provided me with more help than I could possible acknowledge. To their warmth and countless laughters, obrigado, e um cheiro! Resource and funds for the fieldwork were generously provided by Mamdouha Bobst Center for Peace and Justice, the Program for Latin American Studies and the Princeton Institute for International and Regional Studies.

Lastly, my deepest gratitude to my family -- which has now grown! My parents, Sang Ho Kim and Jung Ah Lee, arrived as immigrants in a foreign land, and not only survived, but thrived. Their love, support and guidance throughout my life has led me to where I stand today. 엄마, 아빠, 감사합니다. Carol has been my anchor and the best little sister a brother could ask. I love you so much. And 누나, thank you for your support and kindness, it's wonderful to see how you, Thomas and Nico keep growing.

Finally, to Kimchi and my lifelong partner, Jill Chien. We have made it through adversity and joy, and in the warmth of your embrace, I know that we can become even better, together: kinder, more curious, and pursuant of our hopes and aspirations. I love you all so much, and thank you for always being here. Let's keep moving forward.
