---
title: "The Political Administration: Public Service Delivery and Electoral Accountability in Decentralized Brazil"
author: "Galileu Kim"
date: "`r format(Sys.time(), '%d %B, %Y')`"
documentclass: article
fontsize: 12pt
geometry: [top=1.5in, bottom=1.5in, right=1.5in, left=1.5in]
header-includes:
  - \setlength\parindent{20pt}
  - \usepackage{setspace}
  - \usepackage{multirow}
  - \usepackage{caption}
  - \usepackage{amsmath}
  - \usepackage{setspace}\doublespacing
  - \usepackage{graphicx}
  - \usepackage{subcaption}
  # - \usepackage{floatrow}
  - \DeclareMathOperator*{\argmax}{arg\,max}
  - \pretitle{\centering\Large\textbf}
  - \preauthor{\centering\large}
  - \predate{\centering\normalsize}
bibliography: "`r path.expand('~/princeton/references/dissertation.bib')`"
output: pdf_document
---
```{r setup, include = FALSE}
knitr::opts_chunk$set(
  eval = T,
  cache = T,
  cache.lazy = F,
  message = F,
  warning = F,
  echo = F,
  fig.height = 3.5,
  fig.width = 5,
  fig.align = "center",
  dev = "cairo_pdf"
)

source(
  here::here("scripts/thesis_setup.R")
)
```

# Introduction

```{r child = "chapters/ch0_intro.Rmd"}
```

\pagebreak

# Chapter 1: An electoral theory of political administration

```{r child = "chapters/ch1_model.Rmd"}
```


\pagebreak

# Chapter 2: Going local: public service delivery in education

```{r child = "chapters/ch2_service_delivery.Rmd"}
```


\pagebreak

# Chapter 3: Electoral accountability

```{r child = "chapters/ch3_electoral_accountability.Rmd"}
```


\pagebreak

# Chapter 4: Bureaucratic turnover and educational performance

```{r child = "chapters/ch4_turnover.Rmd"}
```

\pagebreak

# Chapter 5: Coalitional mayoralism

```{r child = "chapters/ch5_coalitional.Rmd"}
```

\pagebreak

# Chapter 6: Beyond politics?

```{r child = "chapters/ch6_beyond_politics.Rmd"}
```

\pagebreak

# Chapter 7: Conclusion

```{r child = "chapters/ch7_conclusion.Rmd"}
```

# References: